---
title: "Toy US election simulator"
output: github_document
---

G. Elliott Morris [\@gelliottmorris](https://www.twitter.com/gelliottmorris)

This is just a simple election simulator based on national and state-level polls. The code in this repo will generate the graphs and statistics I shared [here](https://twitter.com/gelliottmorris/status/1257331350618726400?s=20).

My aim for this code is to help shed some light on basic methods for aggregating national and state polls, inferring electoral standings in states without a lot of data and simulating what might happen in the electoral college if polls lead us astray. **None of this should be considered an official election forecast**, or really even a good one. I bet you'd have better-than-replacement-level rates of success with it, but I only wrote it for a fun coding exercise and to show people how this sort of program works---so act accordingly.

This caveat being addressed, I will concede that I do think this model will provide us with some interesting material as the election cycle progresses, so I've set up the model to update the maps and tables at the bottom of this document throughout the day using [GitHub Actions](https://github.com/features/actions). You can check back here regularly to see how the race is changing.


## Technical notes

The file `scripts/main_poll_simulator.R` runs a series of models to forecast the presidential election using national and state-level polls. The first step is to average available polls fielded over the last two months. That average is weighted by each poll's sample size. If all states had plenty of polls, this model would be easy; we would move on to simulating many different "trial" elections by generating errors from the appropriate distributions. Alas, not all states will be polled adequately, so we turn to an intermediate step.

The second step is to predict what polls would say if pollsters surveyed neglected states. We can regress Biden's observed vote margin in each state on a series of demographic variables in each. I use: Clinton's margin in the 2016 election; the share of adults who are black; the share of adults with a bachelor's degree or higher; the share of adults who are Hispanic or another non-white, non-black race; the median age of voters in a state; the share of adults who are white evangelicals; the average number of people who live within five miles of any given resident; the share of adults who are white and the share of adults who are white without a college degree. Any regular statistical model would struggle to avoid being over-fit by all these variables, so I use [stepwise variable selection via AIC](https://en.wikipedia.org/wiki/Stepwise_regression) and [elastic net regularization](https://en.wikipedia.org/wiki/Elastic_net_regularization) with a linear model trained using [leave-one-out cross-validation](https://www.cs.cmu.edu/~schneide/tut5/node42.html). In states with polls, the predictions from the regression model are given a weight equal to that of a poll with an average sample size and averaged with the raw polling data. In states without polls, the final "polling average" is just the regression prediction.

Because polls are not perfect predictions of voting behavior, the final step is to simulate many tens of thousands of different "trial" election, in each one generating (a) national polling error, (b) a regional polling error and (c) a state-level polling error. These errors are disaggregated from the observed historical root-mean-square error of election polls using a error sum-of-squares [formula](https://fivethirtyeight.com/features/how-fivethirtyeight-calculates-pollster-ratings/) that I cribbed from Nate Silver. This is equivalent to saying that polling error is assumed to be correlated nationally and regionally, but also have state-specific components that aren't shared across geographies. We could be more complex about this—--perhaps someone will submit a pull request to generate correlated state-level errors using `mvrnorm`, for example---but this works for my illustrative purposes here.


## Odds and ends

**A note on timing:** The simulated error in this model is specified to capture the empirical (IE historical) uncertainty in state-level polls fielded **200 days** before the election. It may not be well-calibrated to handle any additional error from the regression model used to "fill in" averages in states without any or many polls. So take it (and the rest of this exercise) as an imperfect guide to the electoral environment, rather than the best or most robust model we could think of.

**A note on forecasting:** The reason this is a "toy" model is because it does not attempt to project movement in the polls between whatever day it runs and election day. Instead, it just treats the polls as uncertain readings of the future, assuming no change in means. But this is an empirically flawed assumption. We know from history that polls during and after conventions tend to over-state the party that most recently nominated a candidate. A true _forecasting_ model will adjust for these historical patterns and project that the favored candidate's election-day polling margin will be smaller than it is on the model run date. This is yet another reason you should treat this analysis with a hefty dose of skepticism---at least until election day...

**A note on polls:** The purpose of this analysis is to determine what we know _now_ from the polls. But polls often err in predicting elections. It is probably better to combine general election polls with other indicators of election outcomes, such as the state of the economy or presidential approval ratings. Fancier election models will do so. This is yet another reason not to squint at the estimates here. 

**A final reminder: this is not an official election forecast.** The purpose of this repo is to help people understand how these forecasts work, and to provide some forecasters with code to improve their methods. 

With all that out of the way, I guess we can proceed...


## Automated report:

![refresh_readme](https://github.com/elliottmorris/toy-us-election-simulator/workflows/refresh_readme/badge.svg)

These graphs are updated hourly with new polls. 

Last updated on **`r format(Sys.time(),'%B %d, %Y at %I:%M %p %Z',tz = 'US/Eastern')`.**


```{r echo=FALSE,message=FALSE,include=FALSE}
source('scripts/main_poll_simulator.R')
```


### National polling average and popular vote prediction

Joe Biden's margin in national polls is **<span style="color: #3498DB;">`r round(og_national_biden_margin*100,1)`</span>** percentage points. 

His margin implied by state-level polls and the demographic regression is **<span style="color: #3498DB;">`r round(national_biden_margin*100,1)`</span>** percentage points.


This chart draws a trend for Biden's implied national margin and plots individual national polls alongside it. His implied national margin will not always match the raw average of national polls.

```{r echo=FALSE,message=FALSE,warning=FALSE}
biden_national_margin_overtime.gg
```


### State polling averages

The polling average in each state:

In map form....

```{r echo=FALSE,message=FALSE,warning=FALSE}
margin_map.gg
```

In table form...

**The twenty most competitive states:**

```{r echo=FALSE,message=FALSE,warning=FALSE}
margin.kable.close
```

**The rest of the states:**

```{r echo=FALSE,message=FALSE,warning=FALSE}
margin.kable.not_close
```


### State polling averages, over time:

_(Just for key states.)_


```{r echo=FALSE,message=FALSE,warning=FALSE,fig.width=8,fig.height=6}
biden_state_margins_overtime.gg
```


### State win probabilities

The odds that either candidate wins a state they're favored in, given the polling error:

In map form...

```{r echo=FALSE,message=FALSE,warning=FALSE,fig.width=8,fig.height=6}
win_probs_map
```


### State win probabilities, over time:

_(Just for key states.)_


```{r echo=FALSE,message=FALSE,warning=FALSE}
biden_state_chance_of_winning_overtime.gg
```


### Tipping-point states

The states that give the winner their 270th electoral college vote, and how often that happens:

```{r echo=FALSE,message=FALSE,warning=FALSE}
tipping_point.kable
```


### Electoral college outcomes

The range of electoral college outcomes:

```{r echo=FALSE,message=FALSE,warning=FALSE}
ev.histogram
```


### Chance of winning the election, over time

```{r echo=FALSE,message=FALSE,warning=FALSE}
biden_chance_of_winning_overtime.gg
```


#### Chance of the popular vote winner losing the electoral college majority

The chance that one party wins the national popular vote, but loses the electoral college majority:

```{r echo = FALSE}
scenarios.kable
```

The probability the Democrats win the national popular vote is `r round(dem_pop_vote_prob*100,2)`%.


#### The divide between the electoral college and popular vote

We can quantify either party's edge as the average across simulations of Joe Biden's margin in the tipping-point state and his margin nationally:

On average, the tipping point state is **<span style="color: #3498DB;">`r abs(round(ev.popvote.divide*100,1))`</span>** percentage points to the **<span style="color: #3498DB;">`r ifelse(ev.popvote.divide < 0,'right','left')`</span>** of the nation as a whole.


### Changes in state averages relative to the national margin

This map shows where Biden and Trump have gained or lost ground since 2016, relative to their gains/losses nationally:

```{r echo=FALSE,message=FALSE,warning=FALSE}
swing.map.gg
```



# Endmatter

I hope you learned something. You can find me on Twitter at [\@gelliottmorris](https://www.twitter.com/gelliottmorris) or my personal website at [gelliottmorris.com](https://www.gelliottmorris.com/).

This content is licensed with the [MIT license](LICENSE).



