---
title: "Toy US election simulator"
output: github_document
---

This is nothing fancy. Just a simple election simulator based on national and state-level polling. The code in this repo will generate the graphs and statistics I shared here: https://twitter.com/gelliottmorris/status/1257331350618726400?s=20

I hope this code well help shed some light on basic methods for aggregating national and state polls, inferring electoral standings in states without a lot of data and simulating what might happen in the electoral college if polls lead us astray. 

**None of this should be considered an official election forecast**, this is just a fun exercise in coding and political statistics.

## Technical notes

The simulated error in this model is specified to capture the empirical (IE historical) uncertainty in state-level polls fielded 200 days before the election. It may not be well-calibrated to handle any additional error from the regression model used to "fill in" averages in states without any or many polls, so take it (and the rest of this exercise) with a grain of salt.

## Automated report:

The following maps and stats are updated periodically throughought the day using [GitHub Actions](https://github.com/features/actions).

Last updated on `r format(Sys.time(),'%B %d, %Y at %H:%M %p')`.




```{r echo=FALSE,message=FALSE,include=FALSE}
source('scripts/main_poll_simulator.R')
```

### National polling average:
Joe Biden's margin in national polls is <span style="color: #3498DB;">`r round(og_national_biden_margin*100,1)`</span> percentage points. That is different than his margin implied by the state-level polls and the demographic regression, which is <span style="color: #3498DB;">`r round(national_biden_margin*100,1)`</span> percentage points.

### State polling averages:
```{r echo=FALSE}
margin_map.gg
```


### Tipping-point state
```{r echo=FALSE}
tipping_point.kable
```

### Electoral college-popular vote divide

On average, the tipping point state is <span style="color: #3498DB;">`r abs(round(ev.popvote.divide*100,1))`</span> percentage points to the <span style="color: #3498DB;">`r ifelse(ev.popvote.divide < 0,'right','left')`</span> of the nation as a whole.

